Variant calling with transformers

### 2022-05-20 manuscript/manuscript.tex
Various manuscript updates

@@

Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$. On each forward pass the model produces two predicted haplotypes. Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls. We record the number of windows in which each variant was detected

@@

\subsection{Genotyping}
 - Are we getting the right variant genotypes

\subsection{Phasing accuracy}
 - look at nearby variants, see if they are phased correctly compared to GIAB information

\subsection{Complex regions}
 - Not sure how to define these, but ideally we'd look for variants in regions where the read mapper had lots of overlapping insertion / deletions in reads and/or softclipping

@@

\subsection{Model size experiments}

 small (5M) medium (10M), large (25M) pretty big (50M)

@@

 \subsection{Phasing accuracy}

@@

We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model. Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task. In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.

 Convolutions - do they make sense?

@@

@@

 \vspace{8pt}
 Conrad, Donald F., et al. "Origins and functional impact of copy number variation in the human genome." Nature 464.7289 (2010): 704.

>>>>

Maybe brief explanation of how transformers work?

@@

Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$. On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position. Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls. For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.

\subsection{Variant quality calculation and false positive removal}

The above variant detection procedure has two shortcomings. First, if every detected variant is emitted, then specificity is poor. Second, variant quality scores are poorly calibrated and do not closely reflect the probability that a variant is real. To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).  The resulting score is used as a final variant quality score that can be used for tuning the tradeoff between sensitivity and specificity.

@@

\subsection{Model size and training data experiments}

We explore the effects of model size and training data size on validation loss and variant detection accuracy. We evaluate three model sizes and two sets of training data.

\begin{center}
	\begin{tabular}{ |c|c|c|c|c| }
	 Model name & Encoder layers & Attention heads & Hidden dimension & Embedding dimension \\ 
	\hline
	 5m & 4 & 4 & 200 & 600 \\ 
	 10m & 6 & 6 & 250 & 400 \\ 
	 25m & 8 & 8 & 400 & 800 \\ 
	 \hline
	\end{tabular}
\end{center}

\subsection{Genotyping \& Phasing}

To check the accuracy of predicted genotypes and phasing of variants we compared calls to the known genotype and, when available, phase from Genome-in-a-Bottle (GIAB). Approximately 99.5\% of homozygous  and 99.0\% of heterozygous calls matched the true genotype, indicating that our model learns to construct both haplotypes accurately.  For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB. The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so our analysis is restricted to only those samples. Approximately 99.4\% of predicted variant pairs had phase that matched the true phase from GIAB.

Accurate phase prediction requires that information regarding which reads support a given haplotype is propogated among input tokens (genomic positions). Our model's relatively high phasing accuracy suggests that

\subsection{Complex regions}
 While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly of reads to identify the true haplotype. We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings in addition to extensive read soft-clipping. 
 To identify such regions we first obtain all indel variants from Genome-in-a-Bottle greater than X bp. We then examine the reads in the alignment file and flag any region that contains more than Y distinct indel start sites or at least Z reads with more than 10bp softclipped. We then

@@

\subsection{Ablation experiments}

@@

@@

We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model. Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish this task. In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.

Our analysis demonstrates that the model learns many features of modern, state of the art variant callers. For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping. In addition, the model learns to employ local context and read-specific features to predict phase

@@

For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work the exploration of the many transformer variations proposed in recent years (vision transformers, swin transformers, fast-former, etc.)

@@
